---
title: "Exploration of the COVID dataset"
output: html_document
---

## Imports
```{r Imports}
library(tidyverse)
library(lubridate)
library(plotrix)
library(vars)
```

## Read Data
```{r Read data}
covid_data = read.csv("other data/covid19-download.csv")
# covid_description = read.csv("other data/covid19-casesdeaths-data-dictionary.csv")
bed = read.csv("top_5_bed_based_shelters.csv")
room = read.csv("top_5_room_based_shelters.csv")
```

## Data Cleaning

To make our shelter data consistent with the date format of the COVID data,
we will group the shelter data into year_week groups (e.g. 202101 means the
first week of 2021).

Furthermore, we are defining Jan 3rd as the start date to be consistent with
the year week aggregation of the COVID data.

```{r Aggregate year week}
aggregate_year_week = function(df) {
  #' Groups df rows in year_week format.
  #' Starts in Jan 3rd 2021 and ends on Sept 21st 2024.
  #' Returns the transformed dataframe.
  #' @param df The dataframe to be transformed
  df %>%
  filter(OCCUPANCY_DATE >= "2021-01-03" & OCCUPANCY_DATE <= "2024-09-21") %>%
  mutate(
    start_of_year = as.Date(paste0(year(OCCUPANCY_DATE), "-01-03")),  # Define Jan 3rd as start
    week_of_year = as.integer(difftime(OCCUPANCY_DATE, start_of_year, units = "days") / 7) + 1,
    year_week = paste0(year(OCCUPANCY_DATE), sprintf("%02d", week_of_year))
  ) %>%
  summarise(
    capacity_actual_sum = sum(capacity_actual_sum),
    capacity_funding_sum = sum(capacity_funding_sum),
    occupied_sum = sum(occupied_sum),
    unoccupied_sum = capacity_actual_sum - occupied_sum,
    unavailable_sum = capacity_funding_sum - capacity_actual_sum,
    occupancy_rate = occupied_sum / capacity_actual_sum,
    .by = c(year_week, shelter_location_pair)
  )
}

# Apply to bed and room dataframes
bed_yr_week = aggregate_year_week(bed)
room_yr_week = aggregate_year_week(room)
```

Next, we want to filter the COVID data to only be within Ontario and within
Jan 3rd 2021 and Sept 21st 2024.

Once that happens, merge the modified_covid_data with bed_yr_week and room_yr_week.

```{r Modify COVID data}
# Add year_week to COVID data
modified_covid_data = covid_data %>%
  filter(date >= "2021-01-03" & date <= "2024-09-21") %>%
  mutate(
    year_week = if_else(
      reporting_week < 10,
      paste0(as.character(reporting_year), "0", as.character(reporting_week)),
      paste0(as.character(reporting_year), as.character(reporting_week))
    )
  ) %>%
  filter(prname == "Ontario")

# Combine room and bed data with COVID data
covid_with_beds = merge(
  x = bed_yr_week,
  y = modified_covid_data,
  by = 'year_week'
)

covid_with_rooms = merge(
  x = room_yr_week,
  y = modified_covid_data,
  by = 'year_week'
)

# Remove original COVID, bed, and room dataframes
rm(covid_data)
rm(bed)
rm(room)
```

Now that we have the shelter dataframes (both bed and room) are merged with
the COVID dataframe respectively, check if the columns are consistent across
both merged dataframes.

```{r Column consistency}
# Make sure all columns are the appropriate datatype
colnames(covid_with_beds) == colnames(covid_with_rooms)
```

All columns are consistent.

Next, clean the data as follows:
- Make NA values consistent (change from "-" to NA)
- Convert dates into proper date objects

```{r}
colnames(covid_with_beds)
```

```{r Clean data}
# Getting numeric columns
# numeric_columns = colnames(select_if(covid_with_rooms, is.numeric))

numeric_columns = list(
  "capacity_actual_sum",
  "capacity_funding_sum",
  "occupied_sum",
  "unoccupied_sum",
  "unavailable_sum",
  "occupancy_rate",
  "totalcases",
  "numtotal_last7",
  "ratecases_total",
  "numdeaths",
  "numdeaths_last7",
  "ratedeaths",
  "ratecases_last7",
  "ratedeaths_last7",
  "numtotal_last14",
  "numdeaths_last14",
  "ratetotal_last14",
  "ratedeaths_last14",
  "avgcases_last7",
  "avgincidence_last7",
  "avgdeaths_last7",
  "avgratedeaths_last7"  
)


# Cleaning numeric columns
for (col in numeric_columns)
{
  # replace -'s with NA
  covid_with_rooms[which(covid_with_rooms[col] == "-"),col] <- NA
  covid_with_beds[which(covid_with_beds[col] == "-"),col] <- NA
  
  # change to numeric datatype
  covid_with_beds[,col] <- as.numeric(covid_with_beds[,col])
  covid_with_rooms[,col] <- as.numeric(covid_with_rooms[,col])
}

# date needs to go from character to date type
covid_with_beds[,"date"] <- as.Date(covid_with_beds[,"date"])
covid_with_rooms[,"date"] <- as.Date(covid_with_rooms[,"date"])
```

## Time series EDA

### ACF/PACF

```{r Plot ACF/PACF function}
plot_ts_acf_pacf = function(df, col, bed_or_room) {
  #' Plots the ACF and PACF of col in df
  #' @param df The dataframe
  #' @param col The column in df
  #' @param bed_or_room Either bed or room
  for (pairing in unique(df$shelter_location_pair)) {
    # filtered_df
    filtered_df = df %>%
      filter(shelter_location_pair == pairing) %>%
      dplyr::select(all_of(col))
  
    # ACF
    acf_result <- acf(
      filtered_df,
      lag.max = 10000,
      plot = FALSE,
      na.action = na.pass
    )
    plot(acf_result, main = paste(
      "ACF of", col, "for", bed_or_room, "based shelter:", pairing
    ))
  
    # PACF
    pacf_result <- pacf(
      filtered_df,
      lag.max = 10000,
      plot = FALSE,
      na.action = na.pass
    )
    plot(pacf_result, main = paste(
      "PACF of", col, "for", bed_or_room, "based shelter:", pairing
    ))
  }
}
```

```{r covid_with_beds$totalcases}
plot_ts_acf_pacf(df = covid_with_beds, col = 'totalcases', bed_or_room = "bed")
```

All ACFs and PACFs for totalcases among the bed shelters are the same since all 5 bed shelters contain 194 year_week records across the data.

```{r covid_with_beds$numdeaths}
plot_ts_acf_pacf(df = covid_with_beds, col = 'numdeaths', bed_or_room = 'bed')
```

All ACFs and PACFs for numdeaths among the bed shelters are the same since all 5 bed shelters contain 194 year_week records across the data.

```{r covid_with_rooms$totalcases}
plot_ts_acf_pacf(df = covid_with_rooms, col = 'totalcases', bed_or_room = 'room')
```

Room shelters 40-1114, 2-1066, and 82-1163 all have all 194 records, leading to the same ACFs and PACFs. Meanwhile, shelter 53-1103 only has 156 records, and shelter 40-1320 only has 133 records.

```{r covid_with_rooms$numdeaths}
plot_ts_acf_pacf(df = covid_with_rooms, col = 'numdeaths', bed_or_room = 'room')
```

Room shelters 40-1114, 2-1066, and 82-1163 all have all 194 records, leading to the same ACFs and PACFs. Meanwhile, shelter 53-1103 only has 156 records, and shelter 40-1320 only has 133 records.

### Scatterplots

In this section, investigate:
- Rate of Total Cases vs Occupancy Rate
- Rate of Deaths vs Occupancy Rate

The investigation will be done per shelter.

```{r Rate of Total Cases vs Occupancy Rate}
# Beds
for (pair in unique(covid_with_beds$shelter_location_pair)) {
    subset = covid_with_beds %>%
      filter(shelter_location_pair == pair) %>%
      dplyr::select(year_week, totalcases, occupancy_rate) %>%
      drop_na()

    plot(
      subset$totalcases,
      subset$occupancy_rate,
      type = 'b',
      main = paste("Rate of Total Cases vs Occupancy Rate for shelter", pair))
}

# Rooms
for (pair in unique(covid_with_rooms$shelter_location_pair)) {
    subset = covid_with_rooms %>%
      filter(shelter_location_pair == pair) %>%
      dplyr::select(year_week, totalcases, occupancy_rate)
    
    plot(
      subset$totalcases,
      subset$occupancy_rate,
      type = 'b',
      main = paste("Rate of Total Cases vs Occupancy Rate for shelter", pair))
}
```

```{r Rate of Deaths vs Occupancy Rate}
# Beds
for (pair in unique(covid_with_beds$shelter_location_pair)) {
    subset = covid_with_beds %>%
      filter(shelter_location_pair == pair) %>%
      dplyr::select(year_week, ratedeaths, occupancy_rate)
    
    plot(
      subset$ratedeaths,
      subset$occupancy_rate,
      type = 'b',
      main = paste("Rate of Deaths vs Occupancy Rate for shelter", pair))
}

# Rooms
for (pair in unique(covid_with_rooms$shelter_location_pair)) {
    subset = covid_with_rooms %>%
      filter(shelter_location_pair == pair) %>%
      dplyr::select(year_week, ratedeaths, occupancy_rate)
    
    plot(
      subset$ratedeaths,
      subset$occupancy_rate,
      type = 'b',
      main = paste("Rate of Deaths vs Occupancy Rate for shelter", pair))
}
```

### Cross-Covariance Function

```{r CCF}
subset = covid_with_beds %>%
  filter(ratedeaths > 120)

ccf(subset$ratedeaths, subset$occupancy_rate, lag.max = 100)
```

### FanChart

```{r VAR and Portmanteau}
x = cbind(subset$occupancy_rate, subset$ratedeaths)
VARselect(x, lag.max = 10, type = "both")
my_fit = VAR(x, p=3, type = 'both')
restrict(my_fit, method = 'ser')

my_residuals <- residuals(my_fit)
plot.ts(my_residuals)

serial.test(my_fit, lags.pt = 16, type = "PT.adjusted")

my_fit_pr <- predict(my_fit, n.ahead = 10, ci = 0.95)
fanchart(my_fit_pr, cis = c(0.95))
```
