---
title: "Extreme Value Analysis"
author: "Erica Fan"
date: "2025-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Steps
- Find outliers (we only really care about lower tail)
- Time series plotting them and observing behaviour (difference between increase and decrease)
- Extreme Value Analysis 
-- (quickly verify fat tailedness from boxplot, hist/density, hypothesis testing?)
-- compute VaR, CVaR, other risk measures
-- inference for tail distribution

## Meta
see EDAnew.Rmd for boxplots

# Find Outliers
```{r libraries}
library(tidyverse)
```

```{r ingest data}
topbed <- read.csv("top_5_bed_based_shelters.csv")
toproom <- read.csv("top_5_room_based_shelters.csv")

topbed$OCCUPANCY_DATE <- as.Date(topbed$OCCUPANCY_DATE)
toproom$OCCUPANCY_DATE <- as.Date(toproom$OCCUPANCY_DATE)
```

```{r}
#boxplot(toproom$occupancy_rate, ylim=c(0.95, 1))
boxplot(toproom$occupancy_rate, main="Top Room Occupancy Rates")
boxplot(topbed$occupancy_rate, main="Top Bed Occupancy Rates")
```


```{r histograms}
# add lines to show Q1 - 1.5*IQR ie. boundary for where outliers begin
q3.bed <- quantile(topbed$occupancy_rate, .75)
q1.bed <- quantile(topbed$occupancy_rate, .25)
q3.room <- quantile(toproom$occupancy_rate, .75)
q1.room <- quantile(toproom$occupancy_rate, .25)

ol.bed <- q1.bed - 1.5*(q3.bed - q1.bed)
ol.room <- q1.room - 1.5*(q3.room - q1.room) 


bks <- seq(0, 1, by=.01)
# top bed
hist(topbed$occupancy_rate, breaks= seq(.6, 1.1, .01))
abline(v=ol.bed)
# top room
hist(toproom$occupancy_rate, breaks=bks)
abline(v=ol.room)

# histogram plot by shelter (par the graphics here)

# plot everything except 100% 
```

Determine our own cutoffs instead of the default boxplot cutoff of $Q1 - 1.5IQR$

Two versions of cutoffs (for now):
* 99%
* Where tail flattens out completely
```{r keep outliers only}

# keep only the outliers

# by default cutoff
outdef.bed <- topbed %>% select(OCCUPANCY_DATE, occupancy_rate) %>% 
  filter(occupancy_rate <= ol.bed)

outdef.room <- toproom %>% select(OCCUPANCY_DATE, occupancy_rate) %>% 
  filter(occupancy_rate <= ol.room)


# with cutoff at 99%
out99.bed <- topbed %>% select(OCCUPANCY_DATE, occupancy_rate) %>% 
  filter(occupancy_rate <= .99)

out99.room <- toproom %>% select(OCCUPANCY_DATE, occupancy_rate) %>% 
  filter(occupancy_rate <= .99)


# with cutoff where it flattens out
outflat.bed <- topbed %>% select(OCCUPANCY_DATE, occupancy_rate) %>% 
  filter(occupancy_rate <= .93)

outflat.room <- toproom %>% select(OCCUPANCY_DATE, occupancy_rate) %>% 
  filter(occupancy_rate <= .97)
```

```{r histogram of outliers only}
#par(mfrow=c(3,2))
hist(outdef.bed$occupancy_rate)
hist(outdef.room$occupancy_rate)
hist(out99.bed$occupancy_rate)
hist(out99.room$occupancy_rate)
hist(outflat.bed$occupancy_rate)
hist(outflat.room$occupancy_rate)
```

```{r outlier ts plots}
ts.plot(outdef.bed$occupancy_rate)
plot(outdef.room$OCCUPANCY_DATE, outdef.room$occupancy_rate, pch=20)
points(outdef.bed$OCCUPANCY_DATE, outdef.bed$occupancy_rate, pch=20, col="red")

plot(out99.room$OCCUPANCY_DATE, out99.room$occupancy_rate, pch=20)
points(out99.bed$OCCUPANCY_DATE, out99.bed$occupancy_rate, pch=20, col="red")

plot(outflat.room$OCCUPANCY_DATE, outflat.room$occupancy_rate, pch=20)
points(outflat.bed$OCCUPANCY_DATE, outflat.bed$occupancy_rate, pch=20, col="red")
```

```{r ts plots but highlighting gaps}

```